---
title: "Hands-On Exercise 1"
format: html
author: "Anabelle Eveleen T"
edtior: visual
date: "2026-04-15"
date-modified: "last-modified"
---

This is a hands on exercise exploration.
```{r}
pacman::p_load(tidyverse)
```
```{r}
exam_data <- read_csv("Exam_data.csv")
```
## Histogram of Exam Data
First, we compare exam data using a histogram
```{r}
hist(exam_data$MATHS)
```
## ggplot objects
```{r}
ggplot(data=exam_data)
```

element into the plot
```{r}
ggplot(data=exam_data, 
       aes(x= MATHS))
```
### geometric objects: 
geom_bar
```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```
geom_dotplot
```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```
geom_histogram
```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram()       
```
we can modify the geometric object by changing geom()
```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light blue")  
```
modify by changing aes()
```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="grey30")
```
geom_density()
```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_density()
```
plotting 2 kernel density lines by using color
```{r}
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()
```
geom_boxplot
```{r}
ggplot(data=exam_data, 
       aes(y = MATHS,       
           x= GENDER)) +    
  geom_boxplot()            
```
notched plot
```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```
geom_violin()
```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()
```
geom_point()
```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()            
```
combination of geom objects
```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)        
```
## statistic function
### stat()
stat_summary()
```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun = "mean",         
               colour ="red",        
               size=4)               
```
combine geom() and stat()
```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",        
             fun="mean",           
             colour="red",          
             size=4)          
```
how to add best fit curve on a scatterplot
```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5)
```
### facets
facet_wrap()
```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```
facet_grid()
```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```
### coordinates
vertical coordinate
```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```
using coord_flip()
```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```
changing the x and y axis range
```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5)
```
```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```
### themes
bar chart using gray color
```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```
using theme_classic()
```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```
using theme_minimal()
```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal() 
```