---
title: "Hands-On Exercise2"
format: html
editor: visual
author: "Anabelle Eveleen T"
date: "2026-04-15"
date-modified: "last-modified"
---

This is a hands on exercise 2 exploration.

In this exercise, ggrepel, ggthemes, hrbrthemes and patchwork will be used for exploration.

```{r}
pacman::p_load(ggrepel, patchwork, 
               ggthemes, hrbrthemes,
               tidyverse) 
```

We will import the data now

```{r}
exam_data <- read_csv("../data/Exam_data.csv", show_col_types = FALSE)
```

## ggrepel

This pacakage is an extension of ggplot2 which provides geoms for ggplot2 to repel overlapping text.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

Here's another example of using ggrepel

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## ggthemes

ggplot2 comes with eight built-in themes. The themes are: theme_gray(), theme_bw(), theme_classic(), theme_dark(), theme_light(), theme_linedraw(), theme_minimal(), and theme_void().

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores") 
```

### ggtheme package: The Economist theme

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```

### ggtheme package: hrbthems

This package provides a base theme that focuses on typographic elements, including where various labels are placed as well as the fonts that are used.

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
```

Utilizing the elements of axis_title_size, base_size and grid to modify the above plot.

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```

## Creating Composite Graphics: pathwork methods

While several ggplot2 extensions---such as gridExtra::grid.arrange() and cowplot::plot_grid()---allow you to combine multiple graphs, the patchwork package offers an intuitive approach. Specifically built for the ggplot2 ecosystem, patchwork simplifies the creation of complex layouts using a syntax that feels like writing a simple equation. General syntax of patchwork are: The Plus Sign (+): Used to place plots side-by-side in a two-column layout. Parentheses (): Used to group subplots together, allowing you to treat multiple graphs as a single unit within a larger layout. The Division Sign (/): Used to stack plots vertically in a two-row layout.

### Combining two ggplot2 graphs

First, we create each single graph naming it p1, p2, p3

```{r}
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")
p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

After creating the graphs, we can utilize th simple patchwork syntax to create a composite of two histograms

```{r}
p1 + p2
```

### Combining three ggplot2 graphs

We aren't just limited to two grids, we can plot more complex composite using the below syntax: The Slash Sign (/): Places plots on top of one another (vertical stacking) The Horizontal Slash (\|): Places plots side-by-side (horizontal stacking) The Parentheses (()): Acts just like the normal parentheses, dictating the order of operations so we can nest complex layouts within a single structure.

```{r}
(p1 / p2) | p3
```

### Creating a composite figure with tag

Patchwork also provides an auto-tagging function to identify subplots in text.

```{r}
((p1 / p2) | p3) + 
  plot_annotation(tag_levels = 'I')
```

### Creating figure with insert

Patchwork offers the inset_element() function.We can define the exact boundaries of your inset using a coordinate system from 0 to 1.

```{r}
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)
```

### Creating a composite figure by using patchwork and ggtheme

We can combine ggthemes and patchwork. Here's an example of a patchwork function and theme_economist()

```{r}
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```
