---
title: "Part B Visualizing and Analysing Text Data"
author: "Anabelle Eveleen T"
format:
  html:
    code-fold: true
    code-summary: "Show code"
    code-tools: true
editor: visual
date: "2026-06-04"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

# Overview

After exploring network data, we will explore how to visualise and analyse text data

By the end of the session, we will be able to:

-   Understand the `tidytext` framework for processing, analysing, and visualising text data.
-   Write functions for importing multiple text files into R.
-   Combine multiple files into a single data frame.
-   Clean and wrangle text data using the `tidyverse` approach.
-   Visualise words with Word Cloud.
-   Compute term frequency-inverse document frequency (TF-IDF) using the `tidytext` method.
-   Visualise text and term relationships.

# Getting Started

## Installing and Launching R Packages

For this exercise, the following R packages will be used for handling, processing, wrangling, analysing, and visualising text data:

-   `tidytext` and `tidyverse` (mainly `readr`, `purrr`, `stringr`, and `ggplot2`) for text processing and general data wrangling.
-   `widyr` for computing pairwise word counts and correlations.
-   `wordcloud` and `ggwordcloud` for generating word cloud visualisations.
-   `textplot` for text-based network plots (requires `igraph`, `tidygraph`, and `ggraph`).
-   `DT` for rendering interactive HTML tables.
-   `lubridate` and `hms` for handling date and time data.

```{r}
pacman::p_load(tidytext, widyr, wordcloud, DT, ggwordcloud, textplot, lubridate, hms,
               tidyverse, tidygraph, ggraph, igraph)
```

# Importing Multiple Text Files from Multiple Folders

## Creating a Folder List

We first define a path pointing to the folder containing the 20 newsgroups dataset.

```{r}
news20 <- "../data/20news/"
```

## Defining a Function to Read All Files from a Folder

The function below reads all files from a given folder and combines them into a single tidy data frame.

```{r}
read_folder <- function(infolder) {
  tibble(file = dir(infolder, 
                    full.names = TRUE)) %>%
    mutate(text = map(file, 
                      read_lines)) %>%
    transmute(id = basename(file), 
              text) %>%
    unnest(text)
}
```

Things to learn from the code chunk:

-   `read_lines()` of the **readr** package is used to read up to n_max lines from a file.
-   `map()` of the **purrr** package is used to transform input by applying a function to each element of a list and returning an object of the same length as the input.
-   `unnest()` of the **tidyr** package is used to flatten a list-column of data frames back out into regular columns.
-   `mutate()` of **dplyr** is used to add new variables while preserving existing ones.
-   `transmute()` of **dplyr** is used to add new variables and drop existing ones.

## Reading in All Messages from the 20news Folder

The code chunk below reads all messages from each newsgroup subfolder and combines them into a single raw text data frame.

```{r}
raw_text <- tibble(folder = 
                     dir(news20, 
                         full.names = TRUE)) %>%
  mutate(folder_out = map(folder, 
                          read_folder)) %>%
  unnest(cols = c(folder_out)) %>%
  transmute(newsgroup = basename(folder), 
            id, text)
write_rds(raw_text, "../data/rds/news20.rds")
```

Things to learn from the code chunk:

-   `dir()` lists all subfolders inside the *news20* path, with full file paths retained using `full.names = TRUE`.
-   `map()` applies `read_folder()` to each subfolder, returning a list of data frames.
-   `unnest()` flattens the list-column of data frames into individual rows.
-   `transmute()` keeps only the *newsgroup*, *id*, and *text* columns for the final output.
-   `write_rds()` saves the combined data frame as an RDS file for future use.

# Initial EDA

The bar chart below shows the frequency of messages by newsgroup, giving an overview of how many individual messages each group contains.

```{r}
raw_text <- read_rds("../data/rds/news20.rds")
raw_text %>%
  group_by(newsgroup) %>%
  summarize(messages = n_distinct(id)) %>%
  ggplot(aes(messages, newsgroup)) +
  geom_col(fill = "lightblue") +
  labs(y = NULL)
```

`group_by(newsgroup)` groups the data by each newsgroup. `summarize(messages = n_distinct(id))` then counts the number of unique message IDs within each group. The resulting summary is passed to `ggplot()` to render a horizontal bar chart, where each bar represents the total message count for one newsgroup.

# Introducing tidytext

The `tidytext` framework applies tidy data principles to text processing, analysis, and visualisation. Much of the infrastructure needed for text mining already exists in packages like `dplyr`, `broom`, `tidyr`, and `ggplot2`, and tidytext is designed to slot into that ecosystem naturally.

## Removing Headers and Automated Email Signatures

Each message in the dataset contains extra structure we do not want to include in our analysis. Every message has a header with fields such as "from:" or "in_reply_to:", and some also have automated email signatures appearing after a line like "--". The code chunk below removes both.

```{r}
cleaned_text <- raw_text %>%
  group_by(newsgroup, id) %>%
  filter(cumsum(text == "") > 0,
         cumsum(str_detect(
           text, "^--")) == 0) %>%
  ungroup()
```

Things to learn from the code chunk:

-   `cumsum()` of base R returns a vector whose elements are the cumulative sums of the elements of the argument. It is used here to identify the position of the first blank line, which marks the end of the header.
-   `str_detect()` from **stringr** is used to detect the presence or absence of a pattern in a string. Here it identifies lines beginning with "--", which signal the start of an email signature.
-   Together, the two `filter()` conditions keep only lines that appear after the first blank line and before any signature separator.

## Removing Nested Quoted Text

In this step, regular expressions are used to remove lines representing quotes from other users, which typically begin with the "\>" character or contain phrasing like "writes:".

```{r}
cleaned_text <- cleaned_text %>%
  filter(str_detect(text, "^[^>]+[A-Za-z\\d]")
         | text == "",
         !str_detect(text,
                     "writes(:|\\.\\.\\.)$"),
         !str_detect(text,
                     "^In article <")
  )
```

Things to learn from the code chunk:

-   `str_detect()` from **stringr** is used to detect the presence or absence of a pattern in a string.
-   The first filter condition retains only lines that do not start with "\>" and contain at least one alphanumeric character, or are blank lines.
-   The second and third conditions remove lines that end with "writes:" or begin with "In article \<", both of which indicate nested quoted content.

## Text Data Processing

`unnest_tokens()` of the **tidytext** package is used to split the dataset into individual word tokens, while `stop_words` is used to filter out common words that carry little analytical meaning.

```{r}
usenet_words <- cleaned_text %>%
  unnest_tokens(word, text) %>%
  filter(str_detect(word, "[a-z']$"),
         !word %in% stop_words$word)
```

With headers, signatures, and formatting removed, we can now explore the most common words across the entire dataset.

```{r}
usenet_words %>%
  count(word, sort = TRUE)
```

Instead of counting individual words across the full dataset, we can also count words within each newsgroup using the code chunk below.

```{r}
words_by_newsgroup <- usenet_words %>%
  count(newsgroup, word, sort = TRUE) %>%
  ungroup()
```

## Visualising Words in Newsgroups

### Static Word Cloud using wordcloud

The code chunk below uses `wordcloud()` of the **wordcloud** package to plot a static word cloud based on word frequencies across all newsgroups.

```{r}
wordcloud(words_by_newsgroup$word,
          words_by_newsgroup$n,
          max.words = 300)
```

`words_by_newsgroup$word` provides the list of words to plot, while `words_by_newsgroup$n` provides the corresponding frequency counts. `max.words = 300` limits the cloud to the top 300 most frequent words, keeping the visualisation readable.

### Interactive Word Cloud using ggwordcloud

The word cloud below is plotted using the **ggwordcloud** package, which integrates with ggplot2 and supports faceting by newsgroup.

```{r}
set.seed(1234)

words_by_newsgroup %>%
  filter(n > 0) %>%
ggplot(aes(label = word,
           size = n)) +
  geom_text_wordcloud() +
  theme_minimal() +
  facet_wrap(~newsgroup)
```

`set.seed(1234)` ensures the word placement is reproducible across runs. `geom_text_wordcloud()` handles the word cloud rendering within each ggplot panel. `facet_wrap(~newsgroup)` splits the word cloud into separate panels for each newsgroup, making it easier to compare which words are prominent in different communities.

# Basic Concept of TF-IDF

TF-IDF (term frequency-inverse document frequency) is a numerical statistic that measures how important a word is to a document within a collection. A word's importance increases with how often it appears in a specific document, but decreases the more frequently it appears across the entire collection. This makes TF-IDF particularly useful for identifying words that are distinctive to individual newsgroups rather than common across all of them.

## Computing TF-IDF within Newsgroups

The code chunk below uses `bind_tf_idf()` of **tidytext** to compute and bind the term frequency, inverse document frequency, and TF-IDF values to the existing dataset.

```{r}
tf_idf <- words_by_newsgroup %>%
  bind_tf_idf(word, newsgroup, n) %>%
  arrange(desc(tf_idf))
```

`bind_tf_idf()` takes the word column, the newsgroup as the document grouping, and the word count *n* as the term frequency input. The function computes all three statistics and appends them as new columns, and `arrange(desc(tf_idf))` sorts the results to surface the most distinctive words first.

## Visualising TF-IDF as an Interactive Table

The code chunk below uses `datatable()` of the **DT** package to create an interactive HTML table that supports pagination, column filtering, and custom formatting.

```{r}
DT::datatable(tf_idf, filter = 'top') %>%
  formatRound(columns = c('tf', 'idf',
                          'tf_idf'),
              digits = 3) %>%
  formatStyle(0,
              target = 'row',
              lineHeight='25%')
```

Things to learn from the code chunk:

-   The `filter = 'top'` argument in `datatable()` adds filter controls above each column, allowing users to narrow down rows interactively.
-   `formatRound()` customises the numeric display of the *tf*, *idf*, and *tf_idf* columns to three decimal places.
-   `formatStyle()` is used to customise the output table. Here, the `target` and `lineHeight` arguments reduce the row line height by 25%, making the table more compact.

## Visualising TF-IDF within Newsgroups

Faceted bar charts are used below to visualise the TF-IDF values of science-related newsgroups, highlighting the words that are most distinctive to each group.

```{r}
tf_idf %>%
  filter(str_detect(newsgroup, "^sci\\.")) %>%
  group_by(newsgroup) %>%
  slice_max(tf_idf,
            n = 12) %>%
  ungroup() %>%
  mutate(word = reorder(word,
                        tf_idf)) %>%
  ggplot(aes(tf_idf,
             word,
             fill = newsgroup)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ newsgroup,
             scales = "free") +
  labs(x = "tf-idf",
       y = NULL)
```

`filter(str_detect(newsgroup, "^sci\\."))` retains only newsgroups whose names begin with "sci.", narrowing the view to science-related communities. `slice_max(tf_idf, n = 12)` selects the top 12 words by TF-IDF score within each group. `mutate(word = reorder(word, tf_idf))` reorders the words so the bars appear in ascending order of TF-IDF within each panel. `facet_wrap(~ newsgroup, scales = "free")` creates a separate panel per newsgroup with independent axis scales, ensuring that differences in word importance are visible even when absolute TF-IDF values differ across groups.

## Counting and Correlating Pairs of Words with widyr

To count the number of times two words appear within the same document, or to measure how correlated they are, the **widyr** package is used. Most pairwise operations require the data to be in wide matrix form first. The widyr workflow casts a tidy dataset into a wide matrix, performs the operation such as a correlation calculation, and then re-tidies the result.

The code chunk below uses `pairwise_cor()` of **widyr** to compute the correlation between newsgroups based on the words they share in common.

```{r}
newsgroup_cors <- words_by_newsgroup %>%
  pairwise_cor(newsgroup,
               word,
               n,
               sort = TRUE)
```

`pairwise_cor()` takes the grouping variable (*newsgroup*), the item variable (*word*), and the count (*n*) to compute pairwise Pearson correlations between all pairs of newsgroups. The `sort = TRUE` argument returns results in descending order of correlation strength.

## Visualising Correlation as a Network

The network graph below visualises the relationships between newsgroups, where edges represent the strength of their word-use correlation.

```{r}
set.seed(2017)

newsgroup_cors %>%
  filter(correlation > .025) %>%
  graph_from_data_frame() %>%
  ggraph(layout = "fr") +
  geom_edge_link(aes(alpha = correlation,
                     width = correlation)) +
  geom_node_point(size = 6,
                  color = "lightblue") +
  geom_node_text(aes(label = name),
                 color = "red",
                 repel = TRUE) +
  theme_void()
```

`filter(correlation > .025)` keeps only newsgroup pairs with a meaningful level of correlation before plotting. `graph_from_data_frame()` from **igraph** converts the filtered data frame into a graph object. `geom_edge_link()` draws edges with both transparency and thickness mapped to the correlation value, so stronger relationships appear as wider, more visible lines. `geom_node_text()` labels each node with the newsgroup name, with `repel = TRUE` preventing label overlap.

## Bigram Analysis

### Creating Bigrams

A bigram is a pair of consecutive words. In this code chunk, `unnest_tokens()` from **tidytext** is used with `token = "ngrams"` and `n = 2` to extract all word pairs from the cleaned text.

```{r}
bigrams <- cleaned_text %>%
  unnest_tokens(bigram,
                text,
                token = "ngrams",
                n = 2)
```

Now, we display the created bigram.

```{r}
bigrams
```

### Counting Bigrams

The code chunk below counts and sorts bigrams in descending order of frequency.

```{r}
bigrams_count <- bigrams %>%
  filter(bigram != 'NA') %>%
  count(bigram, sort = TRUE)
```

```{r}
bigrams_count
```

### Cleaning Bigrams

To remove stop words from bigrams, we first separate each bigram into two individual word columns, then filter out rows where either word is a stop word.

```{r}
bigrams_separated <- bigrams %>%
  filter(bigram != 'NA') %>%
  separate(bigram, c("word1", "word2"),
           sep = " ")

bigrams_filtered <- bigrams_separated %>%
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word)
```

```{r}
bigrams_filtered
```

`separate()` of **tidyr** splits the *bigram* column into two new columns, *word1* and *word2*, using a space as the separator. The two `filter()` calls then remove rows where either word appears in the `stop_words` lexicon, leaving only meaningful word pairs.

### Counting the Cleaned Bigrams

```{r}
bigram_counts <- bigrams_filtered %>%
  count(word1, word2, sort = TRUE)
```

### Creating a Network Graph from the Bigram Data Frame

`graph_from_data_frame()` of the **igraph** package is used to convert the bigram counts into a directed graph object, with each unique word as a node and each bigram pair as a directed edge.

```{r}
bigram_graph <- bigram_counts %>%
  filter(n > 3) %>%
  graph_from_data_frame()
bigram_graph
```

`filter(n > 3)` retains only bigrams that appear more than three times, reducing noise and keeping the graph focused on the most meaningful word associations.

### Visualising a Network of Bigrams with ggraph

The code chunk below uses the **ggraph** package to plot a basic network graph of the bigram relationships.

```{r}
set.seed(1234)

ggraph(bigram_graph, layout = "fr") +
  geom_edge_link() +
  geom_node_point() +
  geom_node_text(aes(label = name),
                 vjust = 1,
                 hjust = 1)
```

### Revised Bigram Network

The revised version below adds directional arrows and maps edge transparency to the bigram count, giving a clearer visual indication of which word pairs appear most frequently.

```{r}
set.seed(1234)

a <- grid::arrow(type = "closed",
                 length = unit(.15,
                               "inches"))

ggraph(bigram_graph,
       layout = "fr") +
  geom_edge_link(aes(edge_alpha = n),
                 show.legend = FALSE,
                 arrow = a,
                 end_cap = circle(.07,
                                  'inches')) +
  geom_node_point(color = "lightblue",
                  size = 5) +
  geom_node_text(aes(label = name),
                 vjust = 1,
                 hjust = 1) +
  theme_void()
```

`grid::arrow(type = "closed", length = unit(.15, "inches"))` creates a closed arrowhead style to indicate the direction of each word pair. `geom_edge_link(aes(edge_alpha = n))` maps the transparency of each edge to the bigram count, so more frequent pairs appear as darker, more prominent connections. `end_cap = circle(.07, 'inches')` adds a small circular buffer at the end of each edge so the arrowhead does not overlap with the node point. `theme_void()` removes all background grid lines and axes for a clean network layout.
