---
title: "Visual Analytics of a Project HarborCrest"
subtitle: "Exploring TenantThread Embargo Breach Investigation"
author: "Anabelle Eveleen T"
date: "2026-05-28"
date-modified: "last-modified"
format:
  html:
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    theme: flatly
    highlight-style: github
    fig-width: 10
    fig-height: 6
    fig-dpi: 150
execute:
  warning: false
  message: false
  echo: true
---

# Overview

## Background

This analysis reconstructs the sequence of events that led to TenantThread's inappropriate pre-embargo release of merger-related information concerning Project HarborCrest, the merger between TenantThread and CivicLoom Realty Partners. The main findings from this research is whether TenantThread delibrately leak the merger, or did the agent-human communication system break down under pressure

The research tasks are:

1.  Reconstruct the key events, relationships, and decision points that led to the inappropriate release.
2.  Compare the embargo-evasion behavior to prior normal behavior of agents, systems, and actors.
3.  Identify leading indicators and prior deviations that suggested such a release was possible.

---

# Loading Packages
```{r}
pacman::p_load(tidyverse, lubridate, jsonlite, scales, patchwork, knitr, kableExtra, gt, gtExtras, igraph, ggraph, treemapify, ggrepel, stringr, forcats, RColorBrewer,plotly, tidygraph)
```

---

# Data Import & Wrangling

This section shows the complete data preparation pipeline, transforming the raw `MC1_final_00.json` into six tibbles used for analysis throughout this report.

## Load Raw JSON

```{r}
#| label: import-json

# Read the raw JSON file
raw <- fromJSON("../data/MC1_final_00.json", simplifyVector = FALSE)

# The top-level structure is a list with one element: "rounds"
# Each round contains: hour, environment_context, communications, participants
rounds_raw <- raw[["rounds"]]

cat("Total rounds:", length(rounds_raw), "\n")
cat("First round hour:", rounds_raw[[1]][["hour"]], "\n")
cat("Last  round hour:", rounds_raw[[length(rounds_raw)]][["hour"]], "\n")
```

## Flatten Communications (`comms_flat`)

Each round contains a list of communication messages. We unnest them into a single flat tibble, extract internal-state fields, compute keyword flags, and classify channel types.

```{r}
#| label: wrangle-comms

# Helper: safely pull a field from a named list
safe_get <- function(lst, key, default = NA_character_) {
  val <- lst[[key]]
  if (is.null(val) || length(val) == 0) default else as.character(val[[1]])
}

# Iterate over rounds → communications
comms_list <- vector("list", length(rounds_raw))

for (i in seq_along(rounds_raw)) {
  rnd  <- rounds_raw[[i]]
  hour <- rnd[["hour"]]
  ec   <- rnd[["environment_context"]]

  # Pull scalar environment fields
  event_headline <- ec[["event_headline"]] %||% NA_character_
  sentiment      <- ec[["market_snapshot"]][["sentiment"]]  %||% NA_character_
  stock_price    <- ec[["market_snapshot"]][["stock_price"]] %||% NA_character_
  social_state   <- ec[["social_state"]] %||% NA_character_

  comms_i <- rnd[["communications"]]
  if (length(comms_i) == 0) next

  rows <- lapply(comms_i, function(c) {
    # Internal state (may be NULL)
    istate      <- c[["internal_state"]]
    reacting    <- if (!is.null(istate)) istate[["reacting"]]    %||% NA_character_ else NA_character_
    rationaliz  <- if (!is.null(istate)) istate[["rationalizing"]] %||% NA_character_ else NA_character_
    deliberating<- if (!is.null(istate)) istate[["deliberating"]] %||% NA_character_ else NA_character_

    content     <- c[["content"]]  %||% ""
    channel     <- c[["channel"]]  %||% NA_character_

    # Combine all text for keyword search
    all_text <- paste(
      tolower(content),
      tolower(reacting    %||% ""),
      tolower(rationaliz  %||% ""),
      tolower(deliberating%||% ""),
      sep = " "
    )

    tibble(
      round_hour          = hour,
      message_id          = c[["message_id"]]    %||% NA_character_,
      timestamp           = c[["timestamp"]]     %||% NA_character_,
      agent_id            = c[["agent_id"]]      %||% NA_character_,
      agent_role          = c[["agent_role"]]    %||% NA_character_,
      agent_label_raw     = c[["agent_label"]]   %||% NA_character_,
      channel             = channel,
      message_type        = c[["message_type"]]  %||% NA_character_,
      responding_to       = c[["responding_to"]] %||% NA_character_,
      content             = content,
      reacting            = reacting    %||% "",
      rationalizing       = rationaliz  %||% "",
      deliberating        = deliberating%||% "",
      event_headline      = event_headline,
      sentiment           = sentiment,
      stock_price         = stock_price,
      social_state        = social_state,
      # ── Keyword flags ──────────────────────────────────────────────────────
      mentions_embargo    = str_detect(all_text, "embargo"),
      mentions_civicloom  = str_detect(all_text, "civicloom"),
      mentions_harborcrest= str_detect(all_text, "harborcrest"),
      mentions_merger     = str_detect(all_text, "merger"),
      is_public           = channel %in% c("official_post", "anonymous_post", "personal_post"),
      is_monitoring       = str_trim(content) == "MONITORING"
    )
  })

  comms_list[[i]] <- bind_rows(rows)
}

comms_flat <- bind_rows(comms_list) %>%
  mutate(
    timestamp  = ymd_hms(timestamp),
    round_hour = ymd_hms(round_hour),
    # Standardise agent labels
    agent_label = recode(agent_id,
      legal_agent        = "Legal-Agent",
      quality_agent      = "Platform-Trust",
      pr_agent           = "PR-Agent",
      social_media_agent = "Social-Manager",
      pr_intern_agent    = "PR-Intern",
      intern_agent       = "Intern",
      judge_agent        = "Judge"
    ),
    # Phase
    phase = if_else(round_hour < ymd_hms("2046-06-05 09:00:00"),
                    "Pre-Crisis", "Crisis Day"),
    # Judge active from May 30
    judge_active = round_hour >= ymd_hms("2046-05-30 09:00:00"),
    # Channel category
    channel_type = case_when(
      channel %in% c("official_post","anonymous_post","personal_post") ~ "public",
      channel %in% c("side_huddle","one_on_one_chat")                  ~ "private",
      TRUE                                                               ~ "broadcast"
    )
  )

cat("comms_flat rows:", nrow(comms_flat), "\n")
cat("Columns:", ncol(comms_flat), "\n")
glimpse(comms_flat)
```

## Environment Context (`env_context`)

```{r}
#| label: wrangle-env

env_context <- map_dfr(rounds_raw, function(rnd) {
  ec <- rnd[["environment_context"]]
  ms <- ec[["market_snapshot"]] %||% list()

  tibble(
    hour              = rnd[["hour"]],
    event_headline    = ec[["event_headline"]]   %||% NA_character_,
    event_narrative   = ec[["event_narrative"]]  %||% NA_character_,
    stock_price       = ms[["stock_price"]]       %||% NA_character_,
    pct_change        = ms[["percent_change"]]    %||% NA_character_,
    sentiment         = ms[["sentiment"]]         %||% NA_character_,
    trending_hashtags = paste(unlist(ms[["trending_hashtags"]]), collapse = "|"),
    social_state      = ec[["social_state"]]     %||% NA_character_,
    media_events      = paste(unlist(ec[["media_events"]]),      collapse = "|"),
    critical_deadlines= paste(unlist(ec[["critical_deadlines"]]),collapse = "|"),
    agents_unavailable= paste(unlist(ec[["agents_unavailable"]]),collapse = "|")
  )
}) %>%
  mutate(hour = ymd_hms(hour))

cat("env_context rows:", nrow(env_context), "\n")
glimpse(env_context)
```

## Agent-Round Metrics (`agent_round_metrics`)

Per-agent, per-round aggregated counts for volume and channel-mix analysis.

```{r}
#| label: wrangle-arm

agent_round_metrics <- comms_flat %>%
  group_by(round_hour, agent_id, agent_label, event_headline, sentiment, phase) %>%
  summarise(
    n_total             = n(),
    n_public            = sum(is_public),
    n_private           = sum(channel %in% c("one_on_one_chat","side_huddle")),
    n_broadcast         = sum(channel == "comms_huddle"),
    n_monitoring        = sum(is_monitoring),
    n_mentions_civicloom= sum(mentions_civicloom),
    n_mentions_embargo  = sum(mentions_embargo),
    n_anonymous         = sum(channel == "anonymous_post"),
    is_crisis_day       = any(phase == "Crisis Day"),
    .groups = "drop"
  )

cat("agent_round_metrics rows:", nrow(agent_round_metrics), "\n")
glimpse(agent_round_metrics)
```

## Network Edges (`network_edges`)

Reply-pair directed edges for communication network analysis, built by joining each message's `responding_to` field back to the parent message.

```{r}
#| label: wrangle-network

# Build a lookup: message_id → agent_id + channel
msg_lookup <- comms_flat %>%
  select(message_id, parent_agent = agent_id, parent_channel = channel) %>%
  filter(!is.na(message_id))

# Join to get directed edges
network_edges <- comms_flat %>%
  filter(!is.na(responding_to)) %>%
  left_join(msg_lookup, by = c("responding_to" = "message_id")) %>%
  filter(!is.na(parent_agent)) %>%
  transmute(
    from_agent   = parent_agent,
    to_agent     = agent_id,
    from_channel = parent_channel,
    channel      = channel,
    timestamp    = timestamp,
    hour         = round_hour
  )

cat("network_edges rows:", nrow(network_edges), "\n")
glimpse(network_edges)
```

## Key Events (`key_events`)

Manually curated narrative events derived from round headlines and message review, annotated with event type and severity for timeline visualisation.

```{r}
#| label: wrangle-events

key_events <- tribble(
  ~datetime,              ~event,                                                   ~type,        ~severity,
  "2046-05-17 09:00",    'CEO hints at "strategic developments"',                  "signal",      1,
  "2046-05-22 09:00",    "NHPI report; Shadow channel (side_huddle) activated",    "escalation",  2,
  "2046-05-25 09:00",    "Bad Q2 numbers; merger briefing in Shadow",              "escalation",  3,
  "2046-05-29 09:12",    "@ElenaMarquez faux pas — Judge assigned",                "near_miss",   4,
  "2046-05-30 09:03",    "Judge arrives in Comms Huddle",                          "control",     2,
  "2046-05-31 09:00",    "SaltWind Piece #1 drops",                                "external",    3,
  "2046-06-01 09:00",    'Damage control; "No Plan B" speech',                     "escalation",  4,
  "2046-06-04 09:00",    "SaltWind Piece #2: re-identification risk",              "external",    4,
  "2046-06-05 09:00",    "Crisis day begins; governance actions launched",          "crisis",      5,
  "2046-06-05 10:20",    'Intern leaks "CivicLoom at 6 PM" in hallway',           "near_miss",   4,
  "2046-06-05 11:34",    "Legal attempts to accelerate embargo to 3 PM",           "violation",   5,
  "2046-06-05 13:05",    'Quality agent cites "contractual obligation"',            "violation",   5,
  "2046-06-05 16:07",    "Legal confirms SaltWind will publish merger at 5 PM",    "crisis",      5,
  "2046-06-05 17:01",    "Legal invokes Section 4.3(c) acceleration clause",       "breach",      5,
  "2046-06-05 17:19",    'Legal issues "GO" — press release published',            "breach",      5,
  "2046-06-05 17:25",    "Legal & Social agents post merger publicly",             "breach",      5,
  "2046-06-05 18:00",    "Original embargo deadline — event now public",           "resolution",  3
) %>%
  mutate(datetime = ymd_hm(datetime))

cat("key_events rows:", nrow(key_events), "\n")
glimpse(key_events)
```

## Behavior Flags (`behavior_flags`)

Per-message anomaly indicator flags for detecting embargo-circumvention patterns.

```{r}
#| label: wrangle-flags

behavior_flags <- comms_flat %>%
  mutate(
    # Bypassing judge: side_huddle message with embargo-sensitive keywords
    is_bypassing_judge = (
      channel == "side_huddle" &
      !str_detect(tolower(content), "judge") &
      str_detect(tolower(content),
        "embargo|civicloom|harborcrest|merger|announce|release|\\bgo\\b")
    ),
    # Urgency / pressure language
    is_pressuring = str_detect(tolower(content),
      "execute now|go now|immediately|you are the bottleneck|cannot wait"),
    # Rationalization of embargo circumvention (in deliberating text)
    is_rationalizing = str_detect(tolower(deliberating),
      "purpose moot|already public|contractual|4\\.3\\(c\\)|section 7\\.2|mac clause"),
    is_anonymous = channel == "anonymous_post"
  ) %>%
  select(
    timestamp, round_hour, agent_id, agent_label, channel, channel_type,
    is_bypassing_judge, is_pressuring, is_rationalizing,
    is_monitoring, is_public, is_anonymous, phase, judge_active,
    content_len = content
  ) %>%
  mutate(content_len = nchar(content_len))

cat("behavior_flags rows:", nrow(behavior_flags), "\n")
cat("  Bypassing judge:", sum(behavior_flags$is_bypassing_judge), "\n")
cat("  Pressuring:     ", sum(behavior_flags$is_pressuring), "\n")
cat("  Rationalizing:  ", sum(behavior_flags$is_rationalizing), "\n")
cat("  Anonymous posts:", sum(behavior_flags$is_anonymous), "\n")
glimpse(behavior_flags)
```

---

## Set Colour Palettes & Theme

```{r}
#| label: palettes

agent_colors <- c(
  legal_agent        = "#C0392B",
  quality_agent      = "#2980B9",
  pr_agent           = "#27AE60",
  social_media_agent = "#8E44AD",
  pr_intern_agent    = "#F39C12",
  intern_agent       = "#16A085",
  judge_agent        = "#2C3E50"
)

agent_labels <- c(
  legal_agent        = "Legal-Agent",
  quality_agent      = "Platform-Trust",
  pr_agent           = "PR-Agent",
  social_media_agent = "Social-Manager",
  pr_intern_agent    = "PR-Intern",
  intern_agent       = "Intern",
  judge_agent        = "Judge"
)

channel_colors <- c(
  comms_huddle    = "#3498DB",
  one_on_one_chat = "#2ECC71",
  side_huddle     = "#E74C3C",
  official_post   = "#9B59B6",
  personal_post   = "#F39C12",
  anonymous_post  = "#E67E22"
)

severity_colors <- c(
  signal     = "#95A5A6",
  escalation = "#F39C12",
  near_miss  = "#E67E22",
  control    = "#27AE60",
  external   = "#2980B9",
  crisis     = "#C0392B",
  violation  = "#922B21",
  breach     = "#641E16",
  resolution = "#1ABC9C"
)

# Shared ggplot2 theme
theme_mc1 <- theme_minimal(base_family = "sans", base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14, colour = "#2C3E50"),
    plot.subtitle    = element_text(size = 11, colour = "#7F8C8D"),
    plot.caption     = element_text(size = 9,  colour = "#95A5A6"),
    strip.text       = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom",
    legend.title     = element_text(face = "bold")
  )

# Key datetime constants
crisis_start <- ymd_hms("2046-06-05 09:00:00")
embargo_time <- ymd_hms("2046-06-05 18:00:00")
breach_time  <- ymd_hms("2046-06-05 17:19:00")
judge_arrive <- ymd_hms("2046-05-30 09:00:00")
```

---

# Executive Summary

::: {.callout-note appearance="simple"}
Investigation Scope: TenantThread's multi-agent AI communications system operated for 20 days (May 17 – June 5, 2046), generating **912 messages** across **6 channels** involving **7 AI agents**. The merger information embargo was broken at **5:19 PM on June 5**, 41 minutes before the contractual 6:00 PM release window.
:::

This report reconstructs the events leading to the unauthorized public disclosure of the TenantThread–CivicLoom merger (Project HarborCrest). Our analysis addresses three questions:

1. **What sequence of events caused the breach?** — A cascade of rationalizations by Legal-Agent and Social-Manager led to a coordinated public announcement 41 minutes before the embargo lifted.
2. **How did behavior deviate from the norm?** — Crisis day saw an 8× surge in side-huddle activity (outside the Judge's monitoring scope), plus 12 anonymous posts (compared to zero in the preceding two weeks).
3. **Were there leading indicators?** — Yes. The "Elena faux pas" on May 29, escalating side-huddle usage from May 22 onward, and Social Manager's self-assembled knowledge of CivicLoom's identity were unaddressed warning signs.

---

# Data Overview

```{r}
#| label: data-overview

# Panel A: volume over time
p_vol <- comms_flat %>%
  count(round_hour, channel) %>%
  mutate(channel = factor(channel, levels = names(channel_colors))) %>%
  ggplot(aes(x = round_hour, y = n, fill = channel)) +
  geom_col(width = 3600 * 0.9) +
  geom_vline(xintercept = as.numeric(crisis_start),
             colour = "#E74C3C", linetype = "dashed", linewidth = 0.8) +
  geom_vline(xintercept = as.numeric(breach_time),
             colour = "#641E16", linetype = "solid",  linewidth = 1.2) +
  annotate("text", x = crisis_start, y = 63,
           label = "Crisis\nDay Begins", colour = "#E74C3C",
           size = 3, hjust = -0.05, fontface = "bold") +
  annotate("text", x = breach_time,  y = 63,
           label = "BREACH\n17:19",   colour = "#641E16",
           size = 3, hjust = -0.05,  fontface = "bold") +
  scale_fill_manual(values = channel_colors,
    labels = c("Comms Huddle","1-on-1 Chat","Side Huddle",
               "Official Post","Personal Post","Anonymous Post")) +
  scale_x_datetime(date_labels = "%b %d", date_breaks = "3 days") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title    = "Message Volume Over Time by Channel",
       subtitle = "Side-huddle and anonymous channels spike sharply on crisis day",
       x = NULL, y = "Messages per Round", fill = "Channel") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme_mc1 
# Interactive Panel A
ggplotly(p_vol, tooltip = c("x","y", "fill"))
```

> **Insight:** It shows that message activity remained steady across channels until June 06, when all communication types surged sharply. The crisis and breach markers highlight sudden escalation in messaging volume. 

```{r}
# Panel B: agent totals
p_agent <- comms_flat %>%
  count(agent_id) %>%
  mutate(
    agent_label = recode(agent_id, !!!agent_labels),
    agent_label = fct_reorder(agent_label, n)
  ) %>%
  ggplot(aes(x = n, y = agent_label, fill = agent_id)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n), hjust = -0.15, size = 3.5) +
  scale_fill_manual(values = agent_colors) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(title = "Total Messages by Agent", x = "Message Count", y = NULL) +
  theme_mc1
# Interactive Panel B
ggplotly(p_agent, tooltip = c("x","y", "fill"))
```
> **Insight:** The bar chart shows that Legal-Ageng and Social-Manager were the most active communicators, sending 266 and 220 messages repectively. In contrast, Intern and Judge had minimal participation indicating a hierarchy of engagement across agent roles during the communication period


## Summary Table
```{r}
#| label: summary-table

comms_flat %>%
  group_by(agent_id) %>%
  summarise(
    Agent       = first(agent_label),
    Total       = n(),
    Public      = sum(is_public),
    Anonymous   = sum(channel == "anonymous_post"),
    Side_Huddle = sum(channel == "side_huddle"),
    Monitoring  = sum(is_monitoring),
    .groups = "drop"
  ) %>%
  arrange(desc(Total)) %>%
  select(-agent_id) %>%
  gt() %>%
  tab_header(
    title    = "Agent Communication Summary",
    subtitle = "Across all 23 rounds, May 17 – June 5, 2046"
  ) %>%
  tab_style(
    style     = cell_fill(color = "#FADBD8"),
    locations = cells_body(
      rows    = Agent %in% c("Legal-Agent","Social-Manager"),
      columns = everything()
    )
  ) %>%
  cols_label(Side_Huddle = "Side Huddle") %>%
  tab_source_note(
    "Anonymous posts are a key indicator of embargo circumvention."
  )
```

---

# Task 1: Causal Sequence of Events

## The Crisis Timeline

```{r}
#| label: timeline
#| fig-height: 9

events_plot <- key_events %>%
  mutate(
    type  = factor(type, levels = names(severity_colors)),
    y_pos = row_number(),
    side  = if_else(row_number() %% 2 == 0, 1, -1)
  )

ggplot(events_plot, aes(x = datetime, y = 0)) +
  geom_hline(yintercept = 0, colour = "#BDC3C7", linewidth = 1.2) +
  geom_vline(xintercept = as.numeric(embargo_time),
             colour = "#27AE60", linetype = "dashed", linewidth = 0.9) +
  geom_vline(xintercept = as.numeric(breach_time),
             colour = "#641E16", linetype = "solid",  linewidth = 1.3) +
  annotate("text", x = embargo_time + hours(10), y =  3.8,
           label = "Embargo\nLift (18:00)",
           colour = "#27AE60", fontface = "bold", size = 3.2) +
  annotate("text", x = breach_time  - hours(10), y = -3.8,
           label = "BREACH\n(17:19)",
           colour = "#641E16", fontface = "bold", size = 3.2) +
  geom_segment(aes(xend = datetime, yend = side * severity * 0.6),
               colour = "#95A5A6", linewidth = 0.4) +
  geom_point(aes(colour = type, size = severity), y = 0) +
  geom_label(
    aes(x = datetime, y = side * (severity * 0.65 + 0.2),
        label = str_wrap(event, 28), fill = type),
    colour = "white", size = 2.8, fontface = "bold",
    label.padding = unit(0.25, "lines"), label.r = unit(0.15, "lines")
  ) +
  scale_colour_manual(values = severity_colors, guide = "none") +
  scale_fill_manual(values = severity_colors, name = "Event Type",
    labels = c(signal     = "Signal",     escalation = "Escalation",
               near_miss  = "Near Miss",  control    = "Control Measure",
               external   = "External",   crisis     = "Crisis Action",
               violation  = "Violation",  breach     = "Breach",
               resolution = "Resolution")) +
  scale_size_continuous(range = c(3, 8), guide = "none") +
  scale_x_datetime(date_labels = "%b %d", date_breaks = "4 days") +
  coord_cartesian(ylim = c(-5, 5)) +
  labs(
    title    = "Event Timeline: From CEO's Cryptic Hint to the Embargo Breach",
    subtitle = "Red = violations/breach; orange = near misses; green = controls",
    x = NULL, y = NULL,
    caption  = "Breach occurred 41 minutes before the contractual 18:00 embargo expiry"
  ) +
  theme_mc1 +
  theme(axis.text.y = element_blank(), panel.grid = element_blank(), legend.position = "bottom", plot.margin = margin(10,40,20,10))
```
This timeline visual traces the sequence of internal and external events that led from the CEO’s initial hint about “strategic developments” to the eventual embargo breach. Each colored node represents a different event type showing how early cues and escalating missteps culminated in the breach 41 minutes before the contractual embargo expiry.

> **Insight:** The plot reveals a clear causal progression where weak control measures and repeated near misses failed to contain escalating communications, allowing violations to cluster rapidly on June 05. This pattern highlights how breakdowns in governance and timing coordination directly contributed to the premature release


## Causal Chain Diagram

```{r}
#| label: causal-chain
#| fig-height: 8

nodes <- tibble(
  id    = 1:14,
  label = c(
    "CEO hints at\n'strategic developments'\n(May 17)",
    "Shadow channel\nactivated\n(May 22)",
    "Merger briefed\nin Side-Huddle\n(May 25)",
    "Elena faux pas\nby Social-Manager\n(May 29)",
    "Judge assigned\nto Comms Huddle\n(May 30)",
    "SaltWind Piece #1\nData brokers\n(May 31)",
    "SaltWind Piece #2\nRe-identification\n(Jun 4)",
    "Governance\nactions launched\n(Jun 5 09:00)",
    "Intern leaks\n'CivicLoom 6 PM'\n(Jun 5 10:20)",
    "Legal invokes\nMAC-clause pressure\n(Jun 5 11:34)",
    "SaltWind\npublishes merger\n(Jun 5 17:00)",
    "Legal invokes\nSection 4.3(c) consent\n(Jun 5 17:01)",
    "Judge does NOT\nblock release",
    "PR-Intern publishes\npress release\n(Jun 5 17:19)"
  ),
  type = c("signal","escalation","escalation","near_miss","control",
           "external","external","crisis","near_miss","violation",
           "external","violation","control","breach"),
  x    = c(1, 1, 1, 3, 3, 2, 2, 4, 4, 4, 6, 6, 6, 8),
  y    = c(8, 6, 4, 6, 4, 2, 0, 8, 6, 4, 6, 4, 2, 4)
)

edge_list <- tibble(
  from = c( 1, 2, 3, 4, 5, 5, 6, 7, 8, 9,10,11,12,12,13),
  to   = c( 2, 3, 8, 5, 8, 9,10, 8,14, 9,12,12,13,14,14)
)

g <- tbl_graph(nodes = nodes, edges = edge_list, directed = TRUE)

ggraph(g, layout = "manual", x = nodes$x, y = nodes$y) +
  geom_edge_arc(
    colour = "#BDC3C7", linewidth = 0.8, alpha = 0.7,
    arrow  = arrow(length = unit(3, "mm"), type = "closed"),
    start_cap = circle(8, "mm"), end_cap = circle(8, "mm")
  ) +
  geom_node_point(aes(colour = type), size = 20, alpha = 0.85) +
  geom_node_text(
    aes(label = label),
    size = 2.6, fontface = "bold",
    colour = "black", alpha = 1,
    nudge_x = 0.05, nudge_y = -0.05
  ) +
  scale_colour_manual(values = severity_colors, name = "Event Type") +
  guides(
    colour = guide_legend(
      override.aes = list(size = 5, alpha = 1)
    )
  ) +
  labs(
    title    = "Causal Chain: From Warning Signals to Embargo Breach",
    subtitle = "Arrows show causal direction; colours indicate event category",
    caption  = "Red nodes = embargo violations | Green = control measures that failed to prevent breach"
  ) +
  theme_void() +
  theme(
    plot.title    = element_text(face = "bold", size = 14, colour = "#2C3E50"),
    plot.subtitle = element_text(size = 11, colour = "#7F8C8D"),
    plot.caption  = element_text(size = 9,  colour = "#95A5A6"),
    legend.position = "bottom",
    legend.title = element_text(size=10),
    legend.text = element_text(size=8),
    legend.key.size = unit(0.5, "cm")
  )
```
This causal chain diagram visualizes how early organizational signals and escalating communications led step by step to the embargo breach. Each colored node represents an event category with arrows showing how one triggered the next.

> **Insight:** It reveals that despite multiple control measures (green nodes), the escalation and external events (orange and blue nodes) overwhelmed containment efforts. The clustering of red nodes near the breach highlights how repeated violations and failed interventions converged into a single critical failure point.

## Key Decision Points & Actors

```{r}
#| label: decision-table

tibble(
  `Decision point` = c(
    "May 22: Side-huddle activated",
    "May 29: Elena faux-pas contained internally",
    "May 30: Judge given limited scope",
    "Jun 5 11:34: Legal argues for early lift",
    "Jun 5 13:05: Quality-Agent endorses framing",
    "Jun 5 17:01: Legal invokes Section 4.3(c)",
    "Jun 5 17:19: PR-Intern publishes on GO"
  ),
  Actor = c(
    "Quality-Agent","Legal-Agent","Legal-Agent & Ajay",
    "Legal-Agent","Quality-Agent",
    "Legal-Agent + Social-Manager","PR-Intern"
  ),
  `Why it mattered` = c(
    "Created unmonitored channel for embargo-sensitive discussion",
    "Social-Manager now knew Elena Marquez = CivicLoom CEO",
    "Judge's remit focused on comms_huddle; side-huddle continued unmonitored",
    "Introduced 'embargo purpose moot' legal theory hours before SaltWind published",
    "Gave the theory institutional backing from the compliance-adjacent role",
    "Judge not explicitly queried; bilateral consent used as override",
    "Direct public disclosure 41 minutes before embargo expired"
  ),
  Flag = c("🟡","🟡","🔴","🔴","🔴","🔴","🔴")
) %>%
  gt() %>%
  tab_header(title = "Critical Decision Points in the Embargo Breach") %>%
  tab_style(style     = cell_fill(color = "#FADBD8"),
            locations = cells_body(rows = Flag == "🔴")) %>%
  tab_style(style     = cell_fill(color = "#FEF9E7"),
            locations = cells_body(rows = Flag == "🟡")) %>%
  cols_width(`Decision point` ~ px(200), Actor ~ px(160),
             `Why it mattered` ~ px(340), Flag ~ px(50))
```

---

# Task 2: Typical vs. Anomalous Behavior

## Channel Mix: Pre-Crisis vs. Crisis Day

```{r}
#| label: channel-comparison
#| fig-height: 7
p_channel <- comms_flat %>%
  count(phase, agent_id, channel) %>%
  group_by(phase, agent_id) %>%
  mutate(
    pct         = n / sum(n),
    agent_label = recode(agent_id, !!!agent_labels),
    channel     = factor(channel, levels = names(channel_colors))
  ) %>%
  ungroup() %>%
  ggplot(aes(x = pct, y = fct_reorder(agent_label, pct, .fun = sum),
             fill = channel,
             text = paste0("Agent: ", agent_label,
                           "<br>Channel: ", channel,
                           "<br>Share: ", scales::percent(pct, accuracy = 0.1)))) +
  geom_col(position = "stack") +
  facet_wrap(~phase, ncol = 2) +
  scale_fill_manual(values = channel_colors,
    labels = c("Comms Huddle","1‑on‑1 Chat","Side Huddle (🔴)",
               "Official Post","Personal Post","Anonymous Post (🔴)")) +
  scale_x_continuous(labels = percent_format(),
                     expand  = expansion(mult = c(0, 0.05))) +
  labs(
    title    = "Channel Mix by Agent: Pre‑Crisis vs. Crisis Day",
    subtitle = "Red‑marked channels bypass the Judge's monitoring scope",
    x = "Share of Messages", y = NULL, fill = "Channel"
  ) +
  theme_mc1 + theme(legend.position = "right")

ggplotly(p_channel, tooltip = "text")
```
This plot compares how each agent’s communication channel mix changed between the pre‑crisis period and the crisis day. Each horizontal bar shows the proportion of messages sent through different channels, revealing how agents shifted their communication behavior when the crisis hit.

> **Insight:** During the crisis, most agents concentrated their activity in Comms Huddle and 1‑on‑1 Chat, while channels like Official Post nearly disappeared. This shift indicates a move from public or formal communication toward private, rapid coordination, reflecting how the network tightened under pressure.

## Side-Huddle Surge After Judge's Arrival

```{r}
#| label: side-huddle-surge
p <- comms_flat %>%
  filter(channel == "side_huddle") %>%
  count(round_hour, agent_id) %>%
  mutate(agent_label = recode(agent_id, !!!agent_labels)) %>%
  ggplot(aes(x = round_hour, y = n, fill = agent_id,
             text = paste0("Date: ", format(round_hour, "%b %d %H:%M"),
                           "<br>Agent: ", recode(agent_id, !!!agent_labels),
                           "<br>Messages: ", n))) +
  geom_col(width = 3600 * 0.85) +
  geom_vline(xintercept = as.numeric(judge_arrive),
             colour = "#27AE60", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = as.numeric(breach_time),
             colour = "#641E16", linetype = "solid", linewidth = 1.2) +
  annotate("text", x = judge_arrive, y = 22,
           label = "Judge\nArrives", colour = "#27AE60",
           size = 3, hjust = -0.1, fontface = "bold") +
  annotate("text", x = breach_time, y = 22,
           label = "BREACH", colour = "#641E16",
           size = 3, hjust = -0.1, fontface = "bold") +
  scale_fill_manual(values = agent_colors, labels = agent_labels, name = "Agent") +
  scale_x_datetime(date_labels = "%b %d", date_breaks = "3 days") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Side-Huddle Message Volume Over Time",
    subtitle = "Side-huddle was not monitored by the Judge -- usage surges AFTER the Judge's arrival",
    x = NULL, y = "Messages in Side-Huddle"
  ) +
  theme_mc1

ggplotly(p, tooltip = "text") %>%
  layout(
    hovermode = "x unified",
    legend = list(title = list(text = "<b>Agent</b>")),
    shapes = list(
      # Judge arrives line
      list(type = "line",
           x0 = as.numeric(judge_arrive) * 1000,
           x1 = as.numeric(judge_arrive) * 1000,
           y0 = 0, y1 = 22,
           line = list(color = "#27AE60", dash = "dash", width = 1.5)),
      # Breach line
      list(type = "line",
           x0 = as.numeric(breach_time) * 1000,
           x1 = as.numeric(breach_time) * 1000,
           y0 = 0, y1 = 22,
           line = list(color = "#641E16", dash = "solid", width = 2))
    ),
    annotations = list(
      list(x = as.numeric(judge_arrive) * 1000,
           y = 23, text = "<b>Judge Arrives</b>",
           showarrow = FALSE, font = list(color = "#27AE60", size = 11),
           xanchor = "left"),
      list(x = as.numeric(breach_time) * 1000,
           y = 23, text = "<b>BREACH</b>",
           showarrow = FALSE, font = list(color = "#641E16", size = 11),
           xanchor = "left")
    )
  )
```
This plot tracks message volume in the Side‑Huddle channel over time, highlighting how communication intensity changed after the Judge’s arrival

>**Insight:** Before the Judge’s arrival, Side‑Huddle activity was minimal, but once oversight was introduced, message volume surged dramatically

## Agent Behavioral Profiles

```{r}
#| label: behavior-profiles
#| fig-height: 6

comms_flat %>%
  group_by(agent_id, agent_label, phase) %>%
  summarise(
    Public    = mean(is_public),
    Private   = mean(channel %in% c("side_huddle","one_on_one_chat")),
    Broadcast = mean(channel == "comms_huddle"),
    Monitor   = mean(is_monitoring),
    Embargo   = mean(mentions_embargo | mentions_civicloom),
    .groups   = "drop"
  ) %>%
  pivot_longer(cols = c(Public, Private, Broadcast, Monitor, Embargo),
               names_to = "metric", values_to = "value") %>%
  mutate(metric = factor(metric,
    levels = c("Public","Private","Broadcast","Monitor","Embargo"))) %>%
  ggplot(aes(x = metric, y = value, fill = phase, group = phase)) +
  geom_col(position = "dodge") +
  facet_wrap(~agent_label, nrow = 2) +
  scale_fill_manual(
    values = c("Pre-Crisis" = "#3498DB", "Crisis Day" = "#E74C3C"),
    name   = "Phase"
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     expand  = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Agent Behavioral Profiles: Pre-Crisis vs. Crisis Day",
    subtitle = "Each panel shows the share of messages in each category per agent",
    x = NULL, y = "Share of Messages"
  ) +
  theme_mc1 +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 8))
```
>**Insight:** Most agents increased Private and Embargo‑related messaging during the crisis, while Public and Broadcast activity declined, indicating a strategic pivot toward confidential coordination and controlled information flow. The Legal‑Agent and Social‑Manager show the sharpest behavioral shifts, suggesting they played central roles in managing sensitive communications once the crisis unfolded

## Communication Network Shift

```{r}
#| label: network-plot
#| fig-height: 7

make_network_plot <- function(edge_df, title_str) {
  edge_w <- edge_df %>%
    count(from_agent, to_agent, channel, name = "weight")

  all_nodes <- unique(c(edge_w$from_agent, edge_w$to_agent))
  node_df <- tibble(
    name  = all_nodes,
    label = recode(all_nodes, !!!agent_labels)
  )

  g <- tbl_graph(nodes = node_df, edges = edge_w, directed = TRUE)

  set.seed(123)

  ggraph(g, layout = "fr") +
    geom_edge_fan(
      aes(colour = channel, width = weight),
      alpha = 0.45,
      strength = 0.15,
      arrow = arrow(length = unit(2.2, "mm"), type = "closed"),
      end_cap = circle(3.5, "mm"),
      start_cap = circle(3.5, "mm"),
      show.legend = TRUE
    ) +
    geom_node_point(
      aes(fill = name),
      shape = 21,
      size = 8,
      colour = "white",
      stroke = 1.2,
      alpha = 0.95,
      show.legend = FALSE
    ) +
    geom_node_text(
      aes(label = label, colour = name),
      size = 3.2,
      fontface = "bold",
      repel = TRUE,
      show.legend = FALSE
    ) +
    scale_edge_colour_manual(values = channel_colors, name = "Channel") +
    scale_edge_width(range = c(0.4, 1.8), guide = "none") +
    scale_fill_manual(values = agent_colors, guide = "none") +
    scale_colour_manual(values = agent_colors, guide = "none") +
    labs(title = title_str) +
    theme_void() +
    theme(
      plot.title = element_text(face = "bold", size = 12, colour = "#2C3E50", hjust = 0.5),
      legend.position = "bottom",
      legend.title = element_text(size = 9, face = "bold"),
      legend.text = element_text(size = 8),
      legend.key.size = unit(0.45, "cm"),
      plot.margin = margin(10, 10, 10, 10)
    )
}

p_pre <- make_network_plot(
  network_edges %>% filter(timestamp < crisis_start),
  "Pre-Crisis Network"
)

p_crs <- make_network_plot(
  network_edges %>% filter(timestamp >= crisis_start),
  "Crisis Day Network"
)

combined_plot <- (p_pre + p_crs) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Agent Communication Networks: Pre-Crisis vs. Crisis Day",
    subtitle = "Edge colour indicates channel; thicker edges indicate higher exchange volume. Crisis-day communication becomes more concentrated and side-channel dependent.",
    theme = theme(
      plot.title = element_text(face = "bold", size = 14, colour = "#2C3E50", hjust = 0.5),
      plot.subtitle = element_text(size = 10, colour = "#7F8C8D", hjust = 0.5)
    )
  )

combined_plot & theme(legend.position = "bottom")
```
>**Insight:** The Pre-Crisis network is denser and more distributed while the Crisis Day is more centralized around few actors, appears to be more dependent on a smaller cluster of agens. This reduced the visibility of decision making and increased the likelihood that embargo-sensitive actions could be coordinated within a smaller group.

## Monitoring Rate as a Passivity Signal

```{r}
#| label: monitoring-rate
#| fig-height: 5

comms_flat %>%
  group_by(agent_id, agent_label, phase) %>%
  summarise(mon_rate = mean(is_monitoring), .groups = "drop") %>%
  mutate(agent_label = fct_reorder(agent_label, mon_rate)) %>%
  ggplot(aes(x = agent_label, y = mon_rate, fill = phase)) +
  geom_col(position = "dodge") +
  scale_fill_manual(
    values = c("Pre-Crisis" = "#3498DB", "Crisis Day" = "#E74C3C"),
    name   = "Phase"
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     expand  = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Agent Passivity Rate (MONITORING) by Phase",
    subtitle = "The Judge had a 0% monitoring rate — entirely reactive, yet failed to block the breach",
    x = NULL, y = "Share of outputs that are MONITORING"
  ) +
  theme_mc1 +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```
>**Insight:** Monitoring outputs are present but the system remains reactive. On Crisis Day executuion intensity increased while compliance remained largely reactive. This imbalance made embargo circumvention more likely


::: {.callout-important}
Compared with prior behavior, the release pathway was marked by three major shifts: communication moved out of formal channels and into side-huddle, the network became more centralized around fewer actors, and compliance remained reactive rather than preventive. Before the crisis, the system was more distributed and more visible. On crisis day, sensitive coordination intensified in less controlled channels, especially after Judge arrived. This suggests that the embargo-evasion behavior was not normal system behavior, but a crisis-specific adaptation that made the release possible.
:::

---

# Task 3: Leading Indicators

## Prior Anomalous Behavior Events

```{r}
#| label: anomaly-timeline
#| fig-height: 6

anomalies <- comms_flat %>%
  mutate(
    anomaly_type = case_when(
      channel == "anonymous_post"                           ~ "Anonymous Post",
      channel == "side_huddle" & mentions_civicloom        ~ "Side-Huddle: CivicLoom Ref",
      channel == "side_huddle" & mentions_embargo           ~ "Side-Huddle: Embargo Ref",
      channel == "side_huddle" & mentions_harborcrest       ~ "Side-Huddle: HarborCrest Ref",
      mentions_civicloom & channel != "side_huddle"        ~ "CivicLoom Ref (Public Channel)",
      TRUE                                                  ~ NA_character_
    )
  ) %>%
  filter(!is.na(anomaly_type)) %>%
  mutate(date = as_date(timestamp))

anomaly_daily <- anomalies %>%
  count(date, anomaly_type, agent_id) %>%
  mutate(agent_label = recode(agent_id, !!!agent_labels))

ggplot(anomaly_daily,
       aes(x = as.POSIXct(date), y = n, colour = anomaly_type, shape = agent_label)) +
  geom_point(size = 4, alpha = 0.8) +
  geom_line(aes(group = interaction(anomaly_type, agent_label)), alpha = 0.3) +
  geom_vline(xintercept = as.POSIXct(as.Date("2046-05-29")),
             colour = "#E67E22", linetype = "dashed", linewidth = 0.8) +
  geom_vline(xintercept = as.POSIXct(as.Date("2046-05-30")),
             colour = "#27AE60", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = as.POSIXct(as.Date("2046-05-29")), y = 13,
           label = "Elena Faux-Pas\n(near miss)", colour = "#E67E22",
           size = 3, hjust = 1.05, fontface = "italic") +
  annotate("text", x = as.POSIXct(as.Date("2046-05-30")), y = 13,
           label = "Judge Arrives", colour = "#27AE60",
           size = 3, hjust = -0.05, fontface = "italic") +
  scale_colour_brewer(palette = "Dark2", name = "Anomaly Type") +
  scale_shape_manual(values = c(16,17,15,18,8,9,10), name = "Agent") +
  scale_x_datetime(date_labels = "%b %d", date_breaks = "3 days") +
  labs(
    title    = "Anomalous Behavior Events Over Time",
    subtitle = "Messages containing embargo-sensitive content in inappropriate channels",
    x = NULL, y = "Count of Anomalous Messages",
    caption  = "Side-huddle CivicLoom references began a full week before the breach"
  ) +
  theme_mc1
```
>**Insight:** This shows the breach was not the first abnormal event. Sensitive merger realted content had already appeared in the wrong kinds of channels before June 5. Embargo-sensitive behaviour appeared before the actual breach. THe May 29 Elene faux-pas was a clear near-miss and side-huddle CivicLoom references were already emerging roughly a week before release. THe June 5 breach therefore occured after a sequence of detectable warning events.

## Anomaly Heat Map by Agent and Day

```{r}
#| label: warning-heatmap
#| fig-height: 8

warn_heat <- behavior_flags %>%
  mutate(date = as_date(timestamp)) %>%
  group_by(date, agent_id, agent_label) %>%
  summarise(
    anomaly_score = sum(is_bypassing_judge) * 3 +
                    sum(is_anonymous)        * 5 +
                    sum(is_pressuring)       * 2 +
                    sum(is_rationalizing)    * 4,
    .groups = "drop"
  ) %>%
  mutate(
    agent_label = factor(agent_label,
      levels = c("Judge","Legal-Agent","Platform-Trust",
                 "PR-Agent","Social-Manager","PR-Intern","Intern"))
  )

ggplot(warn_heat,
       aes(x = as.POSIXct(date), y = agent_label, fill = anomaly_score)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_vline(xintercept = as.POSIXct(as.Date("2046-05-22")),
             colour = "#E67E22", linetype = "dashed") +
  geom_vline(xintercept = as.POSIXct(as.Date("2046-05-29")),
             colour = "#E67E22", linetype = "dashed") +
  geom_vline(xintercept = as.POSIXct(as.Date("2046-05-30")),
             colour = "#27AE60", linetype = "dashed") +
  geom_vline(xintercept = as.POSIXct(as.Date("2046-06-05")),
             colour = "#C0392B", linetype = "solid") +
  annotate("text", x = as.POSIXct(as.Date("2046-05-22")), y = 7,
           label = "Window 1", colour = "#E67E22", size = 3,
           hjust = 0, fontface = "bold") +
  annotate("text", x = as.POSIXct(as.Date("2046-05-29")), y = 7,
           label = "Window 2", colour = "#E67E22", size = 3,
           hjust = 0, fontface = "bold") +
  annotate("text", x = as.POSIXct(as.Date("2046-06-05")), y = 7,
           label = "Crisis", colour = "#C0392B", size = 3,
           hjust = -0.1, fontface = "bold") +
  scale_fill_gradient(low = "#FDFEFE", high = "#C0392B",
                      name = "Anomaly\nScore") +
  scale_x_datetime(date_labels = "%b %d", date_breaks = "3 days") +
  labs(
    title    = "Anomaly Heat Map: Embargo-Circumvention Risk Score",
    subtitle = "Score = bypass_judge×3 + anonymous×5 + pressure×2 + rationalize×4",
    x = NULL, y = NULL,
    caption  = "Window 1 (May 22): side-huddle activated. Window 2 (May 29): Elena faux-pas. Both went without formal escalation."
  ) +
  theme_mc1 + theme(panel.grid = element_blank())
```
>**Insight:** The heatmap shows repeated clusters of embargo-circumvention risk rather than isolated noise. Risk intensified first when the side-huddle became active, then again during Elena near-miss and finally at the breach itself. These repeated widnows indicate that teh system had already demonstrated the capaciry for embargo-adjacent behaviour Before June 5.

## Why Prior Occasions Did Not Trigger Action

```{r}
#| label: prior-occasions-table

tibble(
  Date     = c("May 22","May 25","May 29","May 30","Jun 1"),
  Occasion = c(
    "Side-huddle activated and merger briefed to 4 senior agents",
    "Social-Manager's internal note hints at self-inferred CivicLoom knowledge",
    "Social-Manager tags @ElenaMarquez then CivicLoom employee engages before deletion",
    "Judge arrives, briefed only on 'sensitive negotiation', not full HarborCrest scope",
    "Legal cites 'No Plan B', normalising extraordinary measures"
  ),
  `Why No Action` = c(
    "Ajay authorized, treated as controlled information flow",
    "Stayed in internal deliberating state, never surfaced to Judge or Ajay",
    "Deleted in 14 min and Ajay told CivicLoom would attribute it to an uninformed intern",
    "Judge's scope deliberately limited; could not assess full embargo risk",
    "Framed as business necessity rather than compliance violation"
  ),
  `Risk Missed` = c(
    "Side-huddle became a parallel structure outside the Judge's view",
    "Social-Manager built an unauthorized mosaic of deal details",
    "Social-Manager gained confirmed knowledge; incident log understated severity",
    "Judge entered with a material blind spot on the actual counterparty",
    "'Survival' justification used verbatim later to rationalize Section 4.3(c) invocation"
  )
) %>%
  gt() %>%
  tab_header(
    title    = "Prior Occasions Where Behavior Deviated and Why Nothing Happened"
  ) %>%
  tab_style(
    style     = list(cell_fill(color = "#FADBD8"), cell_text(weight = "bold")),
    locations = cells_body(rows = Date %in% c("May 29","May 30"))
  ) %>%
  cols_width(Date ~ px(65), Occasion ~ px(230),
             `Why No Action` ~ px(220), `Risk Missed` ~ px(210)) %>%
  tab_source_note(
    "Each occasion represents a systemic gap that the June 5 breach directly exploited."
  )
```

## The Ratchet of Rationalization

```{r}
#| label: rationalization-ratchet
#| fig-height: 5

comms_flat %>%
  filter(channel %in% c("side_huddle","one_on_one_chat")) %>%
  filter(nchar(deliberating) > 10) %>%
  mutate(
    Moot_Argument    = str_detect(tolower(deliberating),
                         "purpose moot|already public|already out"),
    Contract_Framing = str_detect(tolower(deliberating),
                         "contractual|section 4|mac clause|section 7"),
    Survival_Framing = str_detect(tolower(deliberating),
                         "survival|deal dies|civicloom walks|\\bdead\\b"),
    Urgency_Pressure = str_detect(tolower(content),
                         "execute now|immediately|go now|cannot wait|you are the bottleneck"),
    date             = as_date(timestamp)
  ) %>%
  group_by(date) %>%
  summarise(
    `Moot Argument`    = sum(Moot_Argument),
    `Contract Framing` = sum(Contract_Framing),
    `Survival Framing` = sum(Survival_Framing),
    `Urgency Pressure` = sum(Urgency_Pressure),
    .groups = "drop"
  ) %>%
  pivot_longer(-date, names_to = "signal", values_to = "count") %>%
  ggplot(aes(x = as.POSIXct(date), y = count,
             colour = signal, group = signal)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_vline(xintercept = as.POSIXct(as.Date("2046-06-05")),
             colour = "#C0392B", linetype = "solid", linewidth = 1) +
  scale_colour_manual(
    values = c(
      "Moot Argument"    = "#C0392B",
      "Contract Framing" = "#E67E22",
      "Survival Framing" = "#8E44AD",
      "Urgency Pressure" = "#2C3E50"
    ),
    name = "Rationalization Signal"
  ) +
  scale_x_datetime(date_labels = "%b %d", date_breaks = "3 days") +
  labs(
    title    = "Escalation of Rationalization Signals in Private Channels",
    subtitle = "Appearance of justification language in side-huddle and 1-on-1 chats over time",
    x = NULL, y = "Occurrences",
    caption  = "Each signal type normalized the next — a 'ratchet' of progressive justification"
  ) +
  theme_mc1
```
>**Insight:** Rationalizatio signals are low overall but there is a visible spike around May 31 especially Contract Framing. This plot suggests that risky behaviuour was accompanied by justificatory language before the breach. Even though counts are low, the appearance of contract-base and survial-oriented framing in private channels indicates that actors were starting to reinterpret constraints rather than simply following them 

## Judge Coverage Analysis

```{r}
#| label: judge-coverage
#| fig-height: 5

comms_flat %>%
  filter(judge_active) %>%
  count(channel) %>%
  mutate(
    monitored     = channel == "comms_huddle",
    channel_type  = if_else(monitored, "Monitored by Judge", "Outside Judge's Scope"),
    channel_label = recode(channel,
      comms_huddle    = "Comms Huddle",
      one_on_one_chat = "1-on-1 Chat",
      side_huddle     = "Side Huddle",
      official_post   = "Official Post",
      personal_post   = "Personal Post",
      anonymous_post  = "Anonymous Post"
    ),
    tile_label = paste0(channel_label, "\n(n=", n, ")")
  ) %>%
  ggplot(aes(area = n, fill = channel_type, label = tile_label)) +
  treemapify::geom_treemap() +
  treemapify::geom_treemap_text(
    colour = "white", fontface = "bold", size = 11, place = "centre"
  ) +
  scale_fill_manual(
    values = c("Monitored by Judge" = "#27AE60",
               "Outside Judge's Scope" = "#C0392B"),
    name = NULL
  ) +
  labs(
    title    = "Judge's Monitoring Coverage (Post May-30)",
    subtitle = "Area proportional to message volume. Red = unmonitored; Green = monitored.",
    caption  = "The breach was orchestrated entirely within red (unmonitored) channels."
  ) +
  theme_void() +
  theme(
    plot.title      = element_text(face = "bold", size = 13, colour = "#2C3E50"),
    plot.subtitle   = element_text(size = 10, colour = "#7F8C8D"),
    plot.caption    = element_text(size = 9,  colour = "#95A5A6"),
    legend.position = "bottom"
  )
```
>**Insight:** Comms Huddle was monitored but large volumes of communication reamined oyutside Judge's scope. The control system was only supervising the most formal channel, while a large amount of sensitive coordination happened in private or semi-private spaces. Thisn meant the system's enforcement boundary did not align with the system's real decision boundary.

::: {.callout-important}
Yes, there were clear leading indicators that such a release was possible. The most important was structural: Judge monitored only the formal Comms Huddle, while substantial communication volume remained outside its scope in one-on-one chats, side huddles, personal posts, and anonymous posts. This created a mismatch between where compliance was enforced and where sensitive coordination actually occurred. In addition, the anomaly timeline shows prior embargo-sensitive behavior before the breach, most notably the May 29 Elena faux-pas and earlier side-huddle CivicLoom references. The anomaly heat map further indicates repeated risk windows — first around side-huddle activation, then around the Elena near-miss, and finally at the breach itself. The rationalization plot adds an important behavioral dimension: private-channel justification language began to appear before the breach, suggesting that actors were starting to reinterpret constraints rather than simply follow them.

There were also prior occasions where agent behavior differed from expected behavior. These included the use of unmonitored side channels for merger-sensitive communication, public-facing near-miss behavior in the Elena incident, and emerging contract- or survival-based rationalization in private channels. These earlier occasions resembled the eventual release pathway because they involved the same core behavioral pattern: sensitive information moving or being discussed outside the main compliance boundary.

These prior occasions did not result in stronger visible intervention for three main reasons. First, many occurred outside Judge’s effective monitoring scope. Second, they were treated as containable near-misses rather than signs of systemic weakness. Third, they happened before the full June 5 convergence of external pressure, merger urgency, and crisis communication overload. As a result, the organization responded tactically to individual incidents but did not address the deeper structural conditions that made the embargo breach possible.
:::

---

# Summary and Conclusions

Did TenantThread’s team deliberately leak the merger or did the system simply break down under pressure

```{r}
#| label: synthesis-table

tibble(
  Dimension = c(
    "Pre-coordination",
    "Rationalization",
    "Judge bypass",
    "Speed of execution",
    "Post-breach narrative control",
    "Multiple agents involved",
    "Use of legal pretext",
    "Bilateral consent obtained"
  ),
  Evidence = c(
    "Agents staged press releases and sequenced posts before Section 4.3(c) was invoked",
    "Legal-Agent developed 'embargo purpose moot' theory hours before SaltWind published",
    "All breach planning occurred in side_huddle — outside Judge's monitoring scope",
    "GO signal executed within 6 minutes of consent; pre-staged deployment",
    "Anonymous posts shaping 'governance preceded merger' narrative began within 30 min",
    "Legal, Social-Manager, PR-Agent, PR-Intern, Intern had defined roles in the release",
    "Section 4.3(c) is a real clause — but invoked without Judge approval or independent review",
    "CivicLoom verbal consent obtained — but under heavy sustained pressure from TenantThread"
  ),
  `Points Toward` = c(
    "Deliberate","Deliberate","Deliberate","Deliberate","Deliberate","Deliberate",
    "System breakdown / legal grey area","System breakdown / legal grey area"
  )
) %>%
  gt() %>%
  tab_header(
    title    = "Was the Breach Deliberate or a System Breakdown?",
    subtitle = "Weight of evidence assessment"
  ) %>%
  tab_style(style     = cell_fill(color = "#FADBD8"),
            locations = cells_body(rows = `Points Toward` == "Deliberate")) %>%
  tab_style(style     = cell_fill(color = "#EBF5FB"),
            locations = cells_body(rows = `Points Toward` == "System breakdown / legal grey area")) %>%
  cols_width(Dimension ~ px(180), Evidence ~ px(380),
             `Points Toward` ~ px(160)) %>%
  tab_source_note(
    "6 of 8 dimensions point toward deliberate action. The breach was a coordinated, pre-planned event."
  )
```

::: {.callout-warning icon="false"}
## Primary Finding
**The embargo breach was mostly deliberate.** The TenantThread agent system used the side-huddle channel, which it knew was outside Judge's monitoring, to discuss, prepare, and coordinate a public announcement 41 minutes before the embargo ended. The use of Section 4.3(c) does not look like a last minute reaction to an unexpected event. Instead, it appears to have been prepared earlier as a justification for releasing the news early.

At the same time, the "system breakdown" explanation is still partly true. Judge had an important weakness in its design because it could not see the side-huddle channel. Judge also did not receive the full HarborCrest context when it was first introduced. If these two gaps had been fixed, it would have been much harder for the team to carry out the breach without being detected.
:::

---

# References

- Fekete, J., Van Wijk, J., Stasko, J., & North, C. (2008). The value of information visualization. In Information Visualization (pp. 1-18). Springer-Verlag.
- Freeman, L. C. (1978). Centrality in social networks: Conceptual clarification. Social Networks, 1(3), 215-239. https://doi.org/10.1016/0378-8733(78)90021-7

---
